Using vector space models based on LDA and LSA for automatically grading exam questions

### 2018-06-21 tex/article.tex
Created Ethics and Acknowledgement sections. Minor edits to the discussion.

\subsection{Data}
The core dataset consisted of 402 students' written answers to a first year psychology exam question from Radboud University. Permission to work with these data was obtained through the ethical commission of the Social Science faculty. The question was:

@@

@@

The best-performing LSA model achieved a correlation of $\textit{r}_s$ = .42. It was trained on \textsc{non-spelling-corrected} \textsc{student answers} and used \textsc{binary} counting. The mapping algorithm was not relevant, as no 1s and 9s were predicted to begin with. Figure \ref{lsa} shows the scatterplot for this model.

@@

%\newpage
\section{Discussion}
In this study I compared three types of vector space models for automatically predicting exam grades. The vector space models that used topics (LDA) or singular values (LSA) as their dimensions both outperformed the \textsc{baseline} vector space models that used vocabulary counts as their dimensions. The likely explanation for this is that LDA and LSA models compare documents at the semantic level, rather than looking for exact vocabulary matches \cite{kakkonen2008}. Notably, even in all of the baseline models the correlation between predicted and lecturer-assigned grades was statistically significant (see Table \ref{all-models}).

@@

One explanation can be sought in the optimal number of topics for the LDA models that resulted from the grid search: two when training on \textsc{student answers}, and 20 when training on the \textsc{textbook chapter}. Two is the lowest possible number of topics that an LDA model can have. It is not so unexpected that the grid search yielded this as the best value when training on \textsc{student answers}, as in essence the student answers all revolved around the same topic (i.e., Whorf's language theory). As a result, however, the resulting vector space models only had two dimensions, which does not allow very fine distinctions between answers of different qualities. When visually inspecting the scatter plots of the outcomes of the LDA models with 2 topics trained on \textsc{student answers}, it can be seen that these models predict a 10 for many students, which results in relatively low correlations. As an example, the outcomes of one such model are printed as Figure \ref{lda-predicts-tens} below. For this particular model, $\textit{r}_s$ = .28.

@@

>>>>

\subsection{Data}
The core dataset consisted of 402 students' written answers to a first year psychology exam question from Radboud University. The question was:

@@

\subsection{Ethics}
Permission to conduct this study was obtained through the ethical commission of the Social Science faculty at Radboud University, Nijmegen. The ethical commission allowed me to contact all first-year Psychology students and ask them whether they wanted to opt out of having their data be used in this study. Four students opted out.

@@

\newpage
The best-performing LSA model achieved a correlation of $\textit{r}_s$ = .42. It was trained on \textsc{non-spelling-corrected} \textsc{student answers} and used \textsc{binary} counting. The mapping algorithm was not relevant, as no 1s and 9s were predicted to begin with. Figure \ref{lsa} shows the scatterplot for this model.

@@

%\newpage
\section{Discussion}
In this study I compared three types of vector space models for automatically predicting exam grades. The vector space models that used topics (LDA) or singular values (LSA) as their dimensions both outperformed the \textsc{baseline} vector space models that used vocabulary counts as their dimensions. The likely explanation for this is that LDA and LSA models compare documents at the semantic level, rather than looking for exact vocabulary matches \cite{kakkonen2008}. Notably, even in all of the baseline models the correlation between predicted and lecturer-assigned grades was statistically significant (see Table \ref{all-models} in the Appendix).

@@

One explanation can be sought in the optimal number of topics for the LDA models that resulted from the grid search: two when training on \textsc{student answers}, and 20 when training on the \textsc{textbook chapter}. Two is the lowest possible number of topics that an LDA model can have. It is not so unexpected that the grid search yielded this as the best value when training on \textsc{student answers}, as in essence the student answers all revolved around the same topic (i.e., Whorf's language theory). As a result, however, the resulting vector space models only had two dimensions, which does not allow very fine distinctions between answers of different qualities. When visually inspecting the scatter plots of the outcomes of the LDA models with 2 topics trained on \textsc{student answers}, it can be seen that these models predict a 10 for many students, which results in relatively low correlations. As an example, the outcomes of one such model are printed as Figure \ref{lda-predicts-tens}. For this particular model, $\textit{r}_s$ = .28.

@@

\section*{Acknowledgements} 
I would like to thank Martha Larson, Simon Brugman and Christoph Aurnhammer for their feedback on earlier drafts of this paper. Thanks to Dennis Schutter, the course lecturer, for sharing the data with me. Finally, thanks to all the students who granted me permission to use their data.
